Neural Connectome of the Ctenophore Statocyst

### 2025-06-24 manuscript/Manuscript.qmd
MS: edits of synapse description

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-system organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems [@Burkhardt_2021]. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour. Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

@@

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2. and not shown). For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 899 skeletonised cells to one of the four quadrants (Figure 1G). <!-- check number of cells, 927 in figure -->

@@

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C). These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. Our results reveal a second type of syncytial neurons in ctenophores (Figure 2--figure supplement C-G).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites (Figure 2C-F). The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F).

@@

Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant. These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 <!--# check --> mitochondria.

@@

Bridge cells had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria. These bridge-cell synapses were formed on ANN neurons or other bridge cells. Bridge cells in the Q1Q2 region formed synapses on ANN_Q1Q2 (3 cells) or on both ANN_Q1Q2 and ANN_Q1-4 (2 cells). Bridge cells in Q3Q4 synapsed on ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).

Nearly all bridge cells (25/26) also received synaptic input from ANNs. Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells). In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells. These cell types and synaptic connections are outside the gravisensory organ and will be described elsewhere.

>>>>

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 972 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-system organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems [@Burkhardt_2021]. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour. Clarifying this relationship between neural architecture and behavioural control is essential to uncovering ctenophore neural function.

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].

A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

@@

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2 and not shown). For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 898 skeletonised AO cells to one of the four quadrants (Figure 1G).

@@

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C). ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. In our dataset, we also reconstructed 10 SNN neurons. These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1). Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement C-G).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites (Figure 2C-F). The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F). SNN neurons had no or only few synapses.

@@

Our synapse annotation revealed several types of synaptic connections within the gravisensory organ. We identified synapses between different ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells and the bridge cells. Bridge cells, but not balancer cells, also synapse on the ANN.

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant. These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 mitochondria.

@@

Nearly all bridge cells (25/26) received synaptic input from ANNs. Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells). In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).

Bridge cells also had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria. These bridge-cell synapses were formed on ANN neurons or other bridge cells. Bridge cells in the Q1Q2 region formed synapses on ANN_Q1Q2 (3 cells) or on both ANN_Q1Q2 and ANN_Q1-4 (2 cells). Bridge cells in Q3Q4 synapsed on ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figures supplement 1). These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.
